
```{r}
# division 240, M210, 240, M220 

test_df <- data.frame(
  species = c(202, 316, 631, 802),
  status = c(1, 1, 0, 1),
  dbh = c(20, 11.1, 11.3, 18.1),
  ht = c(110, 38, 28, 65),
  ha = c(NA, NA, 21, 59),
  broken_top = c("N", "N", "Y", "Y"),
  decay = c(NA, NA, 2, NA),
  cull = c(0, 3, 10, 2)
)

test_df
```

```{r}
EQ_1 <- function(D,H,a,b,c) {
  
  y <- a*(D^b)*(H^c) 
  return(y)
  
}

# k = 9 for softwoods, 11 for hardwoods 
EQ_2 <- function(D,H,k,a,b,b1,c) {
  
  if(D < k) {
    
    y <- a*(D^b)*(H^c)
    
  } else if (D >= k) {
    
    y <- a*(k^(b-b1))*(D^b1)*(H^c) 
    
  }
  
  return(y)
  
}

EQ_3 <- function(D,H,a,a1,b,c,c1) {
  
  y <- a*(D^(a1*(1-exp(-b*D))^c1))*(H^c)
  return(y)
  
}

EQ_4 <- function(D,H,a,b,b1,c) {
  
  y <- a*(D^b)*(H^c)*exp(-(b1*D))
  return(y)
  
}
```


################################################################################
EXAMPLE 1
################################################################################

predict total stem wood volume 
(Table S1)
```{r}
Vtot_ib_Gross <- EQ_2(D = test_df$dbh[1], H = test_df$ht[1], k = 9, a = 0.0019291, b = 2.162413104, b1 = 1.690400253, c = 0.985444005)
Vtot_ib_Gross
```

total stem volumn converted to total stem wood dry weight in lbs 
(specific gravity)
```{r}
Wtot_ib <- Vtot_ib_Gross*0.45*62.4
Wtot_ib
```

reduced weight due to cull
(table 1, harmon et al. 2011) 
```{r}
Wtot_ib_red <- Vtot_ib_Gross*(1 - (test_df$cull[1]/100)*(1-0.54))*0.45*62.4 # I think this should actually be 0.92? 
Wtot_ib_red
```

total stem bark weight 
(Table S6)
```{r}
Wtot_bk <- EQ_1(D = test_df$dbh[1], H = test_df$ht[1], a = 0.009106538, b = 1.437894425, c = 1.336514273)
Wtot_bk
```

total branch weight 
(Table S7)
```{r}
Wbranch <- EQ_1(D = test_df$dbh[1], H = test_df$ht[1], a = 9.521330809, b = 1.762316117, c = -0.405742592)
Wbranch
```

Reductions to bark and branch weights are only considered for dead trees and trees with broken tops
Neither of these conditions is present in the current example 
```{r}
Wtot_bk_red <- Wtot_bk
Wbranchred <- Wbranch
```

Total AGB
(Table S8)
```{r}
AGB_Predicted <- EQ_1(D = test_df$dbh[1], H = test_df$ht[1], a = 0.135206507, b = 1.713527048, c = 1.047613377)
AGB_Predicted 
```

ensure consistent est between the 3 independently estimated components 
```{r}
AGB_Component_red <- Wtot_ib_red + Wtot_bk_red + Wbranchred
AGB_Component_red
```

reduction factor applied to modify AGB_Predicted
```{r}
AGB_Reduce <- AGB_Component_red/(Wtot_ib + Wtot_bk + Wbranch)
AGB_Reduce
```

```{r}
AGB_Predicted_red <- AGB_Predicted*AGB_Reduce
AGB_Predicted_red
```

To harmonize the three components 
```{r}
Wood_Harmonized <- AGB_Predicted_red*(Wtot_ib_red/AGB_Component_red)
Wood_Harmonized 
```

```{r}
Bark_Harmonized <- AGB_Predicted_red*(Wtot_bk_red/AGB_Component_red)
Bark_Harmonized 
```

```{r}
Branch_Harmonized <- AGB_Predicted_red*(Wbranchred/AGB_Component_red)
Branch_Harmonized 
```

predict foliage weight 
(Table S9)
```{r}
Wfoliage <- EQ_2(D = test_df$dbh[1], H = test_df$ht[1], k = 9, a = 0.477184596, b = 2.592670352, b1 = 1.249237429, c = -0.325050455)
Wfoliage
```

reductions in foliage are only considered for live trees with a broken top
as no broken top is present: 
```{r}
Wfoliagered <- Wfoliage
Wfoliagered
```

Carbon content 
(Table 10)
```{r}
C <- AGB_Predicted_red*0.5155958333
C
```


################################################################################
EXAMPLE 2
################################################################################

predict total stem wood volume 
(Table S1)
```{r}
Vtot_ib_Gross <- EQ_1(D = test_df$dbh[2], H = test_df$ht[2], a = 0.001983919, b = 1.810559393, c = 1.129417635)
Vtot_ib_Gross
```

total stem volumn converted to total stem wood dry weight in lbs 
(specific gravity)
```{r}
Wtot_ib <- Vtot_ib_Gross*0.49*62.4
Wtot_ib
```

reduced weight due to cull
(table 1, harmon et al. 2011) 
```{r}
Wtot_ib_red <- Vtot_ib_Gross*(1 - (test_df$cull[2]/100)*(1-0.54))*0.49*62.4
Wtot_ib_red
```

total stem bark weight 
(Table S6)
```{r}
Wtot_bk <- EQ_1(D = test_df$dbh[2], H = test_df$ht[2], a = 0.061595466, b = 1.818642599, c = 0.654020672)
Wtot_bk
Wtot_bk_red <- Wtot_bk
```

total stem weight considering the cull deduction
```{r}
Wtot_ob_red <- Wtot_ib_red + Wtot_bk_red
Wtot_ob_red
```

total branch weight 
(Table S7)
```{r}
Wbranch <- EQ_1(D = test_df$dbh[2], H = test_df$ht[2], a = 0.011144618, b = 3.269520661, c = 0.421304344)
Wbranch
```

For live trees with intact tops, no branch deductions are incurred 
```{r}
Wbranchred <- Wbranch
```

Total AGB
(Table S8)
```{r}
AGB_Predicted <- EQ_4(D = test_df$dbh[2], H = test_df$ht[2], a = 0.315730276, b = 1.853839844, b1 = -0.024745685, c = 0.740557379)
AGB_Predicted 
```

ensure consistent est between the 3 independently estimated components 
```{r}
AGB_Component_red <- Wtot_ib_red + Wtot_bk_red + Wbranchred
AGB_Component_red
```

reduction factor applied to modify AGB_Predicted
```{r}
AGB_Reduce <- AGB_Component_red/(Wtot_ib + Wtot_bk + Wbranch)
AGB_Reduce
```

```{r}
AGB_Predicted_red <- AGB_Predicted*AGB_Reduce
AGB_Predicted_red
```

To harmonize the three components 
```{r}
Wood_Harmonized <- AGB_Predicted_red*(Wtot_ib_red/AGB_Component_red)
Wood_Harmonized 
```

```{r}
Bark_Harmonized <- AGB_Predicted_red*(Wtot_bk_red/AGB_Component_red)
Bark_Harmonized 
```

```{r}
Branch_Harmonized <- AGB_Predicted_red*(Wbranchred/AGB_Component_red)
Branch_Harmonized 
```

predict foliage weight 
(Table S9)
```{r}
Wfoliage <- EQ_1(D = test_df$dbh[2], H = test_df$ht[2], a = 0.850316557, b = 1.99896181, c = -0.418446486)
Wfoliage
```

reductions in foliage are only considered for live trees with a broken top
as no broken top is present: 
```{r}
Wfoliagered <- Wfoliage
Wfoliagered
```

Carbon content 
(Table 10)
```{r}
C <- AGB_Predicted_red*0.4857333333
C
```














