---
output: github_document
---

# UC Berkeley Forest Analytics

The `UCBForestAnalytics` package is a suite of open-source R functions designed to produce standard metrics for forest management and ecology from field inventory data. 


## Installation instructions

To install the `UCBForestAnalytics` package from GitHub:

```r
# install and load devtools
install.packages("devtools")
library(devtools)
```

```r
# install and load Rbiomass
devtools::install_github('kearutherford/UCBForestAnalytics')
library(UCBForestAnalytics)
```


## Tree biomass estimations 

The biomass functions (`TreeBiomass` and `SummaryBiomass`) use Forest Inventory and Analysis (FIA) Regional Biomass Equations to estimate above-ground stem, bark, and branch tree biomass. The functions use the California equation set and should not be used for data from other regions. 


### :eight_spoked_asterisk: `TreeBiomass( )`

The `TreeBiomass` function uses the Forest Inventory and Analysis (FIA) Regional Biomass Equations to estimate above-ground stem, bark, and branch tree biomass. It provides the option to adjust biomass estimates for the structural decay of standing dead trees. Note that standing dead trees also lose mass through degradation, which is indirectly accounted for in this method through a decrease in height. However, degradation of bark and branches is not directly accounted for. See references 1, 2 & 3 below. 

### Inputs

1. `data` A dataframe or tibble. Each row must be an observation of an individual tree.

2. `status` Must be a variable (column) in the provided dataframe or tibble. Specifies whether the individual tree is alive (1) or dead (0). The class of this variable will be coerced to factor.

3. `species` Must be a variable (column) in the provided dataframe or tibble. Specifies the species of the individual tree. Must follow four-letter species code or FIA naming conventions (see species code tables in background information). The class of this variable will be coerced to character.

4. `dbh` Must be a **numeric** variable (column) in the provided dataframe or tibble. Provides the diameter at breast height (DBH) of the individual tree in either centimeters or inches. 

5. `ht` Must be a **numeric** variable (column) in the provided dataframe or tibble. Provides the height of the individual tree in either meters or feet.

6. `decay_class` Default is set to "ignore", indicating that biomass estimates for standing dead trees will not be adjusted for structural decay. It can be set to a variable (column) in the provided dataframe or tibble. For standing dead trees, the decay class should be 1, 2, 3, 4, or 5 (see decay class descriptions in background information). For live trees, the decay class should be NA or 0. The class of this variable will be coerced to character.

7. `sp_codes` Not a variable (column) in the provided dataframe or tibble. Specifies whether the species variable follows the four-letter code or FIA naming convention (see species code tables below). Must be set to either "4letter" or "fia". The default is set to "4letter".

8. `units` Not a variable (column) in the provided dataframe or tibble. Specifies whether the dbh and ht variables were measured using metric (centimeters and meters) or imperial (inches and feet) units. Also specifies whether the results will be given in metric (kilograms) or imperial (US tons) units. Must be set to either "metric" or "imperial". The default is set to "metric".

### Outputs

The original dataframe will be returned, with four new columns. If decay_class is provided, the biomass estimates for standing dead trees will be adjusted for structural decay.

1. `stem_bio_kg` (or `stem_bio_tons`): biomass of stem in kilograms (or US tons)

2. `bark_bio_kg` (or `bark_bio_tons`): biomass of bark in kilograms (or US tons)

3. `branch_bio_kg` (or `branch_bio_tons`): biomass of branches in kilograms (or US tons)

4. `total_bio_kg` (or `total_bio_tons`): biomass of tree (stem + bark + branches) in kilograms (or US tons)

*Important note: For some hardwood species, the `stem_bio` includes bark and branch biomass. In these cases, bark and branch biomass are not available as separate components of total biomass. `bark_bio` and `branch_bio` will appear as `NA` and the `total_bio` will be equivalent to the `stem_bio`.*

### Demonstrations

```{r, include = FALSE}
library(UCBForestAnalytics)
```

```{r}
# investigate input dataframe
bio_demo_data
```

```{r}
# call the TreeBiomass() function in the UCBForestAnalytics package
# keep default decay_class (= "ignore"), sp_codes (= "4letter") and units (= "metric")
tree_bio_demo1 <- TreeBiomass(data = bio_demo_data,
                              status = "Live",
                              species = "SPP",
                              dbh = "DBH_CM",
                              ht = "HT_M")

tree_bio_demo1
```

**Notice in the output dataframe:**

* QUKE (California black oak) has `NA` `bark_bio_kg` and `branch_bio_kg`. For some hardwood species, the `stem_bio_kg` includes bark and branch biomass. In these cases, bark and branch biomass are not available as separate components of total biomass. 

* The column names of the input dataframe will remain intact in the output dataframe.

* The `Forest`, `Plot_id`, `SPH`, and `Decay` columns, which are not directly used in the biomass calculations, remain in the output dataframe. Any additional columns in the input dataframe will remain in the output dataframe.

<br>

```{r}
# call the TreeBiomass() function in the UCBForestAnalytics package
# keep default decay_class (= "ignore"), sp_codes (= "4letter") and units (= "metric")
tree_bio_demo2 <- TreeBiomass(data = bio_demo_data,
                              status = "Live",
                              species = "SPP",
                              dbh = "DBH_CM",
                              ht = "HT_M",
                              decay_class = "Decay",
                              sp_codes = "4letter",
                              units = "metric")

tree_bio_demo2
```

**Notice in the output dataframe:**

* Comparing between the outputs from demo1 and demo2:
    * For the three standing dead trees, the biomass estimates are adjusted for structural decay. 
    * For the live trees, the biomass estimates remain the same. 

<br>

### :eight_spoked_asterisk: `SummaryBiomass( )`

The `SummaryBiomass` function calls on the `TreeBiomass` function described above. Additionally, the outputs are summarized by plot or by plot as well as species. 

### Inputs

1. `data` A dataframe or tibble. Each row must be an observation of an individual tree.

2. `site` Must be a variable (column) in the provided dataframe or tibble. Describes the broader location or forest where the data were collected. The class of this variable will be coerced to character.

3. `plot` Must be a variable (column) in the provided dataframe or tibble. Identifies the plot in which the individual tree was measured. The class of this variable will be coerced to character.

4. `exp_factor` Must be a numeric variable (column) in the provided dataframe or tibble. The expansion factor specifies the number of trees per hectare (or per acre) that a given plot tree represents.

5. `status` Must be a variable (column) in the provided dataframe or tibble. Specifies whether the individual tree is alive (1) or dead (0). The class of this variable will be coerced to factor.

6. `decay_class` Must be a variable (column) in the provided dataframe or tibble. For standing dead trees, the decay class should be 1, 2, 3, 4, or 5 (see decay class descriptions in background information). For live trees, the decay class should be NA or 0. The class of this variable will be coerced to character.

6. `species` Must be a variable (column) in the provided dataframe or tibble. Specifies the species of the individual tree. Must follow four-letter species code or FIA naming conventions (see species code tables in background information). The class of this variable will be coerced to character.

7. `dbh` Must be a **numeric** variable (column) in the provided dataframe or tibble. Provides the diameter at breast height (DBH) of the individual tree in either centimeters or inches. 

8. `ht` Must be a **numeric** variable (column) in the provided dataframe or tibble. Provides the height of the individual tree in either meters or feet.

9. `sp_codes` Not a variable (column) in the provided dataframe or tibble. Specifies whether the species variable follows the four-letter code or FIA naming convention (see species code tables below). Must be set to either "4letter" or "fia". The default is set to "4letter".

10. `units` Not a variable (column) in the provided dataframe or tibble. Specifies (1) whether the dbh and ht variables were measured using metric (centimeters and meters) or imperial (inches and feet) units; (2) whether the expansion factor is in metric (stems per hectare) or imperial (stems per acre) units; and (3) whether results will be given in metric (megagrams per hectare) or imperial (US tons per acre) units. Must be set to either "metric" or "imperial". The default is set to "metric".

11. `results` Not a variable (column) in the provided dataframe or tibble. Specifies whether the results will be summarized by plot or by plot as well as species. Must be set to either "by_plot" or "by_species." The default is set to "by_plot".

### Outputs

A dataframe with the following columns: 

1. `site`: as described above

2. `plot`: as described above

3. `species`: if results argument was set to "by_species"

4. `live_Mg_ha` (or `live_ton_ac`): above-ground live tree biomass in megagrams per hectare (or US tons per acre)

5. `dead_Mg_ha` (or `dead_ton_ac`): above-ground dead tree biomass in megagrams per hectare (or US tons per acre)

### Demonstrations 

```{r}
# investigate input dataframe
bio_demo_data
```

<br>

**Results summarized by plot:**
```{r}
# call the SummaryBiomass() function in the UCBForestAnalytics package
# keep default sp_codes (= "4letter") and units (= "metric")
sum_bio_demo1 <- SummaryBiomass(data = bio_demo_data,
                                site = "Forest",
                                plot = "Plot_id",
                                exp_factor = "SPH",
                                status = "Live",
                                decay_class = "Decay",
                                species = "SPP",
                                dbh = "DBH_CM",
                                ht = "HT_M",
                                results = "by_plot")

sum_bio_demo1
```

<br>

**Results summarized by plot as well as by species:**
```{r}
# call the SummaryBiomass() function in the UCBForestAnalytics package
# keep default sp_codes (= "4letter") and units (= "metric")
sum_bio_demo2 <- SummaryBiomass(data = bio_demo_data,
                                site = "Forest",
                                plot = "Plot_id",
                                exp_factor = "SPH",
                                status = "Live",
                                decay_class = "Decay",
                                species = "SPP",
                                dbh = "DBH_CM",
                                ht = "HT_M",
                                results = "by_species")

sum_bio_demo2
```


<br>

**If there are plots without trees:**
```{r}
# investigate input dataframe
bio_NT_demo
```

```{r}
# call the SummaryBiomass() function in the UCBForestAnalystics package
sum_bio_demo3 <- SummaryBiomass(data = bio_NT_demo,
                                site = "Forest",
                                plot = "Plot_id",
                                exp_factor = "SPH",
                                status = "Live",
                                decay_class = "Decay",
                                species = "SPP",
                                dbh = "DBH_CM",
                                ht = "HT_M",
                                results = "by_plot")

sum_bio_demo3
```

*Notice that the plot without trees has 0 live and dead biomass.*


<br>

## Forest composition and structure compilations  

The forest composition and structure functions (`ForestComp` and `ForestStr`) assist with common plot-level data compilations. These functions help ensure that best practices in data compilation are observed.


### :eight_spoked_asterisk: `ForestComp( )`

### Inputs

1. `data` A dataframe or tibble. Each row must be an observation of an individual tree.

2. `site` Must be a variable (column) in the provided dataframe or tibble. Describes the broader location or forest where the data were collected. The class of this variable will be coerced to character.

3. `plot` Must be a variable (column) in the provided dataframe or tibble. Identifies the plot in which the individual tree was measured. The class of this variable will be coerced to character.

4. `exp_factor` Must be a **numeric** variable (column) in the provided dataframe or tibble. The expansion factor specifies the number of trees per hectare (or per acre) that a given plot tree represents.

5. `status` Must be a variable (column) in the provided dataframe or tibble. Specifies whether the individual tree is alive (1) or dead (0). The class of this variable will be coerced to factor.

6. `species` Must be a variable (column) in the provided dataframe or tibble. Specifies the species of the individual tree. The class of this variable will be coerced to character.

7. `dbh` Must be a **numeric** variable (column) in the provided dataframe or tibble. Provides the diameter at breast height (DBH) of the individual tree in either centimeters or inches.

8. `relative` Not a variable (column) in the provided dataframe or tibble. Specifies whether forest composition should be measured as relative basal area or relative density. Must be set to either "BA" or "density". The default is set to "BA".

9. `units` Not a variable (column) in the provided dataframe or tibble. Specifies whether the dbh variable was measured using metric (centimeters) or imperial (inches) units. Must be set to either "metric" or "imperial". The default is set to "metric".

### Outputs

A dataframe with the following columns: 

1. `site`: as described above

2. `plot`: as described above

3. `species`: as described above 

4. `dominance`: relative basal area (or relative density) in percent (%). Only compiled for LIVE trees. 

### Demonstrations

```{r}
# investigate input dataframe
for_demo_data
```

<br>

**Composition measured as relative basal area:**
```{r}
# call the ForestComp() function in the UCBForestAnalystics package
# keep default relative (= "BA") and units (= "metric")
comp_demo1 <- ForestComp(data = for_demo_data,
                         site = "Forest",
                         plot = "Plot_id",
                         exp_factor = "SPH",
                         status = "Live",
                         species = "SPP",
                         dbh = "DBH_CM")

comp_demo1
```

<br>

**Composition measured as relative density:**
```{r}
# call the ForestComp() function in the UCBForestAnalystics package
comp_demo2 <- ForestComp(data = for_demo_data,
                         site = "Forest",
                         plot = "Plot_id",
                         exp_factor = "SPH",
                         status = "Live",
                         species = "SPP",
                         dbh = "DBH_CM",
                         relative = "density",
                         units = "metric")

comp_demo2
```

<br>

**If there are plots without trees:**
```{r}
# investigate input dataframe
for_NT_demo
```

```{r}
# call the ForestComp() function in the UCBForestAnalystics package
comp_demo3 <- ForestComp(data = for_NT_demo,
                         site = "Forest",
                         plot = "Plot_id",
                         exp_factor = "SPH",
                         status = "Live",
                         species = "SPP",
                         dbh = "DBH_CM")

comp_demo3
```

*Notice that the plot without trees has NA dominance for all species.*

<br>

### :eight_spoked_asterisk: `ForestStr( )`

### Inputs

1. `data` A dataframe or tibble. Each row must be an observation of an individual tree.

2. `site` Must be a variable (column) in the provided dataframe or tibble. Describes the broader location or forest where the data were collected. The class of this variable will be coerced to character.

3. `plot` Must be a variable (column) in the provided dataframe or tibble. Identifies the plot in which the individual tree was measured. The class of this variable will be coerced to character.

4. `exp_factor` Must be a **numeric** variable (column) in the provided dataframe or tibble. The expansion factor specifies the number of trees per hectare (or per acre) that a given plot tree represents.

5. `dbh` Must be a **numeric** variable (column) in the provided dataframe or tibble. Provides the diameter at breast height (DBH) of the individual tree in either centimeters or inches.

6. `ht` Default is set to "ignore", which indicates that tree heights were not taken. If heights were taken, it can be set to a **numeric** variable (column) in the provided dataframe or tibble, providing the height of the individual tree in either meters or feet.

7. `units` Not a variable (column) in the provided dataframe or tibble. Specifies (1) whether the dbh and ht variables were measured using metric (centimeters and meters) or imperial (inches and feet) units; (2) whether the expansion factor is in metric (stems per hectare) or imperial (stems per acre) units; and (3) whether results will be given in metric or imperial units. Must be set to either "metric" or "imperial". The default is set to "metric".

### Outputs

A dataframe with the following columns: 

1. `site`: as described above

2. `plot`: as described above

3. `sph` (or `spa`): stems per hectare (or stems per acre)

4. `ba_m2_ha` (or `ba_ft2_ac`): basal area in meters squared per hectare (or feet squared per acre)

5. `qmd_cm` (or `qmd_in`): quadratic mean diameter in centimeters (or inches). Weighted by the expansion factor.

6. `dbh_cm` (or `dbh_in`): average diameter at breast hegiht in centimeters (or inches). Weighted by the expansion factor.

7. `ht_m` (or `ht_ft`): average height in meters (or feet) if ht argument was set. Weighted by the expansion factor.

### Demonstrations

```{r}
# investigate input dataframe
for_demo_data
```

<br>

**If tree heights were not measured:**
```{r}
# call the ForestStr() function in the UCBForestAnalystics package
# keep default ht (= "ignore") and units (= "metric")
str_demo1 <- ForestStr(data = for_demo_data,
                       site = "Forest",
                       plot = "Plot_id",
                       exp_factor = "SPH",
                       dbh = "DBH_CM")

str_demo1
```

<br>

**If tree heights were measured:**
```{r}
# call the ForestStr() function in the UCBForestAnalystics package
str_demo2 <- ForestStr(data = for_demo_data,
                       site = "Forest",
                       plot = "Plot_id",
                       exp_factor = "SPH",
                       dbh = "DBH_CM",
                       ht = "HT_M",
                       units = "metric")

str_demo2
```

<br>

**If there are plots without trees:**
```{r}
# investigate input dataframe
for_NT_demo
```

```{r}
# call the ForestStr() function in the UCBForestAnalystics package
str_demo3 <- ForestStr(data = for_NT_demo,
                       site = "Forest",
                       plot = "Plot_id",
                       exp_factor = "SPH",
                       dbh = "DBH_CM",
                       ht = "HT_M",
                       units = "metric")

str_demo3
```

*Notice that the plot without trees has 0 stems/ha, 0 basal area, NA QMD, NA DBH, and NA height.*

<br>

## Surface and ground fuel load estimations 

The three functions (`FineFuels`, `CoarseFuels` and `GroundFuels`) estimate surface and ground fuel loads from line-intercept transects. Field data should have been collected following Brown (1974) or a similar method. 

This set of functions evolved from Rfuels, a package developed by Danny Foster ([See Rfuels GitHub](https://github.com/danfosterfire/Rfuels)). 
Although these functions are formatted differently than Rfuels, they follow the same general equations. The goal of this set of functions is to take the workflow outlined in Rfuels and make it more flexible and user-friendly. Rfuels will remain operational as the legacy program. 


### :eight_spoked_asterisk: `FineFuels( )`

The `FineFuels` function estimates fine woody debris (FWD) loads. FWD is defined as 1-hour (0-0.64cm or 0-0.25in), 10-hour (0.64-2.54cm or 0.25-1.0in), and 100-hour (2.54-7.62cm or 1-3in) fuels. Assumptions for FWD data collection: 

* Each plot has one or more fuel transects (the number of transects per plot is variable)
* For pre-defined distances along each transect, the number of 1-hour, 10-hour, and 100-hour fuel particles that intersect the transect are recorded as counts
* The distances (or lengths) sampled along the transect may be different for each size class (1-hour, 10-hour, and 100-hour)

### Inputs

1. `tree_data` A dataframe or tibble. Each row must be an observation of an individual tree. Must have at least these columns (column names are exact):

    * **time:** Describes the time period of measurement (could be the year, the month, etc. depending on the project). The class of this variable will be coerced to character.
    * **site:** Describes the broader location or forest where the data were collected. The class of this variable will be coerced to character.
    * **plot:** Identifies the plot in which the individual tree was measured. The class of this variable will be coerced to character.
    * **exp_factor:** The expansion factor specifies the number of trees per hectare (or per acre) that a given plot tree represents. The class of this variable must be numeric. 
    * **species:** Specifies the species of the individual tree. Must follow four-letter species code or FIA naming conventions (see species code tables in background information). The class of this variable will be coerced to character. 
    * **dbh:** Provides diameter at breast height of the individual tree in either centimeters or inches. The class of this variable must be numeric.

2. `fuel_data` A dataframe or tibble. Each row must be an observation of an individual transect at a specific time/site/plot. Must have at least these columns (column names exact):

    * **time:** Described the time period of measurement (could be the year, the month, etc. depending on the project). The class of this variable will be coerced to character.
    * **site:** Describes the broader location or forest where the data were collected. The class of this variable will be coerced to character.
    * **plot:** Identifies the plot in which the individual fuel transect was measured. The class of this variable will be coerced to character.
    * **transect:** Identifies the transect on which the specific fuel tallies were collected. The transect ID Will often be an azimuth from plot center. The class of this variable will be coerced to character.
    * **count_1h:** Transect counts of the number of intersections for 1-hour fuels. Must be an integer greater than or equal to 0. 
    * **count_10h:** Transect counts of the number of intersections for 10-hour fuels. Must be an integer greater than or equal to 0.
    * **count_100h:** Transect counts of the number of intersections for 100-hour fuels. Must be an integer greater than or equal to 0.
    * **length_1h:** The length of the sampling transect for 1-hour fuels in either meters or feet. The class of this variables must be numeric. 
    * **length_10h:** The length of the sampling transect for 10-hour fuels in either meters or feet. The class of this variables must be numeric. 
    * **length_100h:** The length of the sampling transect for 100-hour fuels in either meters or feet. The class of this variables must be numeric. 
    * **slope:** The slope of the transect in percent (not the slope of the plot). This column is OPTIONAL. However, it is important to correct for the slope effect on the horizontal length of transects. If slope is not supplied, the slope will be taken to be 0 (no slope). 

3. `sp_codes` Specifies whether the species column in tree_data follows the four-letter code or FIA naming convention (see species code tables below). Must be set to either “4letter” or “fia”. The default is set to “4letter”.

4. `units` Specifies whether the input data are in metric (centimeters, meters, and trees per hectare) or imperial (inches, feet, and trees per acre) units. Inputs must be all metric or all imperial (do not mix-and-match units). The output units will match the input units (i.e., if inputs are in metric then outputs will be in metric). Must be set to either "metric" or "imperial". The default is set to "metric".

*Note: there must be a one-to-one match between time:site:plot identities of tree and fuel data.*

### Outputs

A dataframe with the following columns: 

1. `time`: as described above

2. `site`: as described above 

3. `plot`: as described above 

4. `load_1h_Mg_ha` (or `load_1h_ton_ac`): fuel load of 1-hour fuels in megagrams per hectare (or US tons per acre)

5. `load_10h_Mg_ha` (or `load_10h_ton_ac`): fuel load of 10-hour fuels in megagrams per hectare (or US tons per acre)

6. `load_100h_Mg_ha` (or `load_100h_ton_ac`): fuel load of 100-hour fuels in megagrams per hectare (or US tons per acre)

7. `load_fwd_Mg_ha` (or `load_fwd_ton_ac`): total fine woody debris fuel load (1-hour + 10-hour + 100-hour) in megagrams per hectare (or US tons per acre)


### Demonstration

```{r}
# investigate input tree_data
overstory_demo
```

```{r}
# invesigate input fuel_data 
fwd_demo
```

<br>

```{r}
# call the FineFuels() function in the UCBForestAnalystics package
# keep default sp_codes (= "4letter") and units (= "metric")
fine_demo <- FineFuels(tree_data = overstory_demo,
                       fuel_data = fwd_demo)

fine_demo
```

<br>

### :eight_spoked_asterisk: `CoarseFuels( )`

The `CoarseFuels` function estimates coarse woody debris (CWD) loads. CWD is defined 1000-hour (7.62+ cm or 3+ in) fuels. Assumptions for CWD data collection: 

* Each plot has one or more fuel transects (the number of transects per plot is variable)
* For a pre-defined distance along each transect, every 1000-hour fuel particle that intersects the transect is individually recorded (diameter and decay status)


### Inputs

1. `tree_data` A dataframe or tibble. Each row must be an observation of an individual tree. Must have at least these columns (column names are exact):

    * **time:** Describes the time period of measurement (could be the year, the month, etc. depending on the project). The class of this variable will be coerced to character.
    * **site:** Describes the broader location or forest where the data were collected. The class of this variable will be coerced to character.
    * **plot:** Identifies the plot in which the individual tree was measured. The class of this variable will be coerced to character.
    * **exp_factor:** The expansion factor specifies the number of trees per hectare (or per acre) that a given plot tree represents. The class of this variable must be numeric. 
    * **species:** Specifies the species of the individual tree. Must follow four-letter species code or FIA naming conventions (see species code tables in background information). The class of this variable will be coerced to character. 
    * **dbh:** Provides diameter at breast height of the individual tree in either centimeters or inches. The class of this variable must be numeric.

2. `fuel_data` A dataframe or tibble with at least these columns (column names exact):

    * **time:** Described the time period of measurement (could be the year, the month, etc. depending on the project). The class of this variable will be coerced to character.
    * **site:** Describes the broader location or forest where the data were collected. The class of this variable will be coerced to character.
    * **plot:** Identifies the plot in which the individual fuel transect was measured. The class of this variable will be coerced to character.
    * **transect:** Identifies the transect on which the specific fuel tallies were collected. The transect ID Will often be an azimuth from plot center. The class of this variable will be coerced to character.
    * **length_1000h:** The length of the sampling transect for 1000-hour fuels in either meters or feet. The class of this variable must be numeric. 
    * **slope:** The slope of the transect in percent (not the slope of the plot). This column is OPTIONAL. However, it is important to correct for the slope effect on the horizontal length of transects. If slope is not supplied, the slope will be taken to be 0 (no slope). 
    
    * If sum-of-squared-diameters for sound and rotten 1000-hour fuels has already been calculated by the user, the dataframe must also have the following two columns. In this case, each row is an observation of an individual transect at a specific time/site/plot.
        * **ssd_S:** Sum-of-squared-diameters for sound 1000-hour fuels. The class of this variable must be numeric.
        * **ssd_R:** Sum-of-squared-diameters for rotten 1000-hour fuels. The class of this variable must be numeric.
    
    * If sum-of-squared-diameters for sound and rotten 1000-hour fuels has NOT already been calculated by the user, the dataframe must also have the following two columns. In this case, each row is an observation of an individual 1000-hour fuel particle recorded at a specific time/site/plot/transect. 
        * **diameter:** Diameter of the individual 1000-hour fuel particle in either centimeters or inches.The class of this variable must be numeric. 
        * **status:** Decay status of the individual 1000-hour fuel particle. Must be either "R" (rotten) or "S" (sound). The class of this variable will be coerced to character.

3. `sp_codes` Specifies whether the species column in tree_data follows the four-letter code or FIA naming convention (see species code tables below). Must be set to either “4letter” or “fia”. The default is set to “4letter”.

4. `units` Specifies whether the input data are in metric (centimeters, meters, and trees per hectare) or imperial (inches, feet, and trees per acre) units. Inputs must be all metric or all imperial (do not mix-and-match units). The output units will match the input units (i.e., if inputs are in metric then outputs will be in metric). Must be set to either "metric" or "imperial". The default is set to "metric".

5. `summed` Specifies whether the sum-of-squared-diameters for sound and rotten 1000-hour fuels has already been calculated by the user. Must be set to either "yes" or "no". The default is set to "no".

*Note: there must be a one-to-one match between time:site:plot identities of tree and fuel data.*

### Outputs

A dataframe with the following columns: 

1. `time`: as described above

2. `site`: as described above 

3. `plot`: as described above 

4. `load_1000s_Mg_ha` (or `load_1000s_ton_ac`): fuel load of sound 1000-hour fuels in megagrams per hectare (or US tons per acre)

5. `load_1000r_Mg_ha` (or `load_1000r_ton_ac`): fuel load of rotten 1000-hour fuels in megagrams per hectare (or US tons per acre)

6. `load_cwd_Mg_ha` (or `load_cwd_ton_ac`): total coarse woody debris fuel load (1000-hour sound + 1000-hour rotten) in megagrams per hectare (or US tons per acre)


### Demonstrations

```{r}
# investigate input tree_data
overstory_demo
```

<br> 

**If sum-of-squared-diameters for sound and rotten 1000-hour fuels has already been calculated:**
```{r}
# invesigate input fuel_data 
cwd_YS_demo
```

<br>

```{r}
# call the CoarseFuels() function in the UCBForestAnalystics package
coarse_demo1 <- CoarseFuels(tree_data = overstory_demo,
                            fuel_data = cwd_YS_demo,
                            sp_codes = "4letter",
                            units = "metric",
                            summed = "yes")

coarse_demo1
```

<br> 

**If sum-of-squared-diameters for sound and rotten 1000-hour fuels has NOT already been calculated:**
```{r}
# invesigate input fuel_data 
cwd_NS_demo
```

*Notice that time:site:plot:transects without fuels are represented with a diameter of 0 and an NA status. Status could also be set to either "S" or "R". It is important that transects without CWD are still included, as those transects indicate a loading of 0.*

<br>

```{r}
# call the CoarseFuels() function in the UCBForestAnalystics package
# keep default sp_codes (= "4letter"), units (= "metric"), and summed (= "no")
coarse_demo2 <- CoarseFuels(tree_data = overstory_demo,
                            fuel_data = cwd_NS_demo)

coarse_demo2
```

<br>

### :eight_spoked_asterisk: `GroundFuels( )`

The `GroundFuels` function estimates duff and litter loads. Assumptions for duff/litter data collection: 

* Each plot has one or more fuel transects (the number of transects per plot is variable)
* Duff and litter depths (or combined duff/litter depth) are sampled at fixed locations along each transect (the number of samples per transect are variable) 


### Inputs

1. `tree_data` A dataframe or tibble. Each row must be an observation of an individual tree. Must have at least these columns (column names are exact):

    * **time:** Describes the time period of measurement (could be the year, the month, etc. depending on the project). The class of this variable will be coerced to character.
    * **site:** Describes the broader location or forest where the data were collected. The class of this variable will be coerced to character.
    * **plot:** Identifies the plot in which the individual tree was measured. The class of this variable will be coerced to character.
    * **exp_factor:** The expansion factor specifies the number of trees per hectare (or per acre) that a given plot tree represents. The class of this variable must be numeric. 
    * **species:** Specifies the species of the individual tree. Must follow four-letter species code or FIA naming conventions (see species code tables in background information). The class of this variable will be coerced to character. 
    * **dbh:** Provides diameter at breast height of the individual tree in either centimeters or inches. The class of this variable must be numeric.

2. `fuel_data` A dataframe or tibble with at least these columns (column names exact):

    * **time:** Described the time period of measurement (could be the year, the month, etc. depending on the project). The class of this variable will be coerced to character.
    * **site:** Describes the broader location or forest where the data were collected. The class of this variable will be coerced to character.
    * **plot:** Identifies the plot in which the individual fuel transect was measured. The class of this variable will be coerced to character.
    * **transect:** Identifies the transect on which the specific fuel tallies were collected. The transect ID Will often be an azimuth from plot center. The class of this variable will be coerced to character.
        
    * If duff and litter depth are measured separately, the dataframe must also have the following two columns:
        * **litter_depth:** Litter depth in centimeters or inches. May be an individual depth measurement or the average depth on the transect (see note below). The class of this variable must be numeric.
        * **duff_depth:** Duff depth in centimeters or inches. May be an individual depth measurement or the average depth on the transect (see note below). The class of this variable must be numeric.
    
    * If duff and litter depth are measured together, the dataframe must also have the following column: 
        * **lit_duff_depth:** Combined litter and duff depth in centimeters or inches. May be an individual depth measurement or the average depth on the transect (see note below). The class of this variable must be numeric.
    
    *Note: If multiple depth measurements were taken for each transect, the user may average the depths together before import (in which case each row is an observation of an individual transect at a specific time/site/plot) or not average the depths before import (in which case each row is an observation of an individual depth recorded at a specific time/site/plot/transect).* 

3. `sp_codes` Specifies whether the species column in tree_data follows the four-letter code or FIA naming convention (see species code tables below). Must be set to either “4letter” or “fia”. The default is set to “4letter”.

4. `units` Specifies whether the input data are in metric (centimeters, meters, and trees per hectare) or imperial (inches, feet, and trees per acre) units. Inputs must be all metric or all imperial (do not mix-and-match units). The output units will match the input units (i.e., if inputs are in metric then outputs will be in metric). Must be set to either "metric" or "imperial". The default is set to "metric".

5. `measurement` Specifies whether duff and litter were measured together or separately. Must be set to "combined" or "separate". The default is set to "separate".

*Note: there must be a one-to-one match between time:site:plot identities of tree and fuel data.*

### Outputs

A dataframe with the following columns: 

1. `time`: as described above

2. `site`: as described above 

3. `plot`: as described above 

    *If duff and litter were measured separately:*

    * `litter_Mg_ha` (or `litter_ton_ac`): litter load in megagrams per hectare (or US tons per acre)

    * `duff_Mg_ha` (or `duff_ton_ac`): duff load in megagrams per hectare (or US tons per acre)

    *If duff and litter were measured together:*

    * `lit_duff_Mg_ha` (or `lit_duff_ton_ac`): combined litter and duff load in megagrams per hectare (or US tons per acre)


### Demonstrations

```{r}
# investigate input tree_data
overstory_demo
```

<br> 

**If depths were NOT averaged together for each transect before import:**
```{r}
# invesigate input fuel_data 
lit_duff_demo
```

<br>

```{r}
# call the GroundFuels() function in the UCBForestAnalystics package
# keep default sp_codes (= "4letter"), units (= "metric"), and measurement (= "separate")
duff_demo1 <- GroundFuels(tree_data = overstory_demo,
                          fuel_data = lit_duff_demo)

duff_demo1
```

<br> 

**If depths were averaged together for each transect before import:**
```{r}
# invesigate input fuel_data 
lit_duff_avg_demo
```

<br>

```{r}
# call the GroundFuels() function in the UCBForestAnalystics package
duff_demo2 <- GroundFuels(tree_data = overstory_demo,
                          fuel_data = lit_duff_avg_demo,
                          sp_codes = "4letter",
                          units = "metric",
                          measurement = "separate")

duff_demo2
```

<br>

## Background information for tree biomass estimations 

### Species code tables 

**Softwoods**

|common name|scientific name|4-letter code|FIA code|
|:---|:---|:---|:---|
|White fir|Abies concolor|ABCO|15|
|Grand fir|Abies grandis|ABGR|17|
|California red fir|Abies magnifica|ABMA|20|
|Noble fir|Abies procera|ABPR|22|
|Western juniper|Juniperus occidentalis|JUOC|64|
|Incense cedar|Calocedrus decurrens|CADE|81|
|Lodgepole pine|Pinus contorta|PICO|108|
|Jeffrey pine|Pinus jeffreyi|PIJE|116|
|Sugar pine|Pinus lambertinana|PILA|117|
|Western white pine|Pinus monticola|PIMO|119|
|Ponderosa pine|Pinus ponderosa|PIPO|122|
|Foothill pine|Pinus sabiniana|PISA|127|
|Douglas-fir|Pseudotsuga menziesii|PSME|202|
|Redwood|Sequoioideae sempervirens|SESE|211|
|Giant sequoia|Sequoiadendron giganteum|SEGI|212|
|Pacific yew|Taxus brevifolia|TABR|231|
|California nutmeg|Torreya californica|TOCA|251|
|Western hemlock|Tsuga heterophylla|TSHE|263|
|Mountain hemlock|Tsuga mertensiana|TSME|264|
|Unknown conifer|NA|UNCO|298|

<br>

**Hardwoods** 

|common name|scientific name|4-letter code|FIA code|
|:---|:---|:---|:---|
|Bigleaf maple|Acer macrophyllum|ACMA|312|
|White alder|Alnus rhombifolia|ALRH|352|
|Pacific madrone|Arbutus menziesii|ARME|361|
|Golden chinkapin|Chrysolepis chrysophylla|CHCH|431|
|Pacific dogwood|Cornus nuttallii|CONU|492|
|Tanoak|Notholithocarpus densiflorus|NODE|631|
|Quaking aspen|Populus tremuloides|POTR|746|
|California live oak|Quercus agrifolia|QUAG|801|
|Canyon live oak|Quercus chrysolepis|QUCH|805|
|California black oak|Quercus kelloggii|QUKE|818|
|Willow species|Salix spp.|SASP|920|
|California-laurel|Umbellularia californica|UMCA|981|
|Unknown hardwood|NA|UNHA|998|
|Unknown tree|NA|UNTR|999|

*Note: Four-letter species codes are the first two letters of the genus followed by the first two letters of the species.*

### FIA standing dead tree decay class descriptions 

|Decay class|Limbs and branches|Top|% bark remaining|Sapwood presence and condition|Heartwood condition|
|:---|:---|:---|:---|:---|:---|
|1|All present|Pointed|100|Intact; sound, incipient decay, hard, original color|Sound, hard, original color|
|2|Few limbs, no fine branches|May be broken|Variable|Sloughing; advanced decay, fibrous, firm to soft, light brown|Sound at base, incipient decay in outer edge of upper bole, hard, light to reddish brown|
|3|Limb studs only|Broken|Variable|Sloughing; fibrous, soft, light to reddish brown|Incipient decay at base, advanced decay throughout upper bole, fibrous, hard to firm, reddish brown|
|4|Few or no studs|Broken|Variable|Sloughing; cubical, soft, reddish to dark crown|Advanced decay at base, sloughing from upper bole, fibrous to cubical, soft, dark reddish brown|
|5|None|Broken|Less than 20|Gone|Sloughing, cubical, soft, dark brown, OR fibrous, very soft, dark reddish brown, encased in hardened shell|

### References 

**For Regional Biomass Equations**

* *Volume Estimation for the PNW-FIA Integrated Database.* USDA Forest Service, Pacific Northwest Research Station. [Link to PDF](https://ww2.arb.ca.gov/sites/default/files/cap-and-trade/protocols/usforest/2014/volume_equations.pdf)

* *Regional Biomass Equations Used by FIA to Estimate Bole, Bark, and Branches.* USDA Forest Service, Pacific Northwest Research Station. [Link to PDF](https://ww2.arb.ca.gov/sites/default/files/cap-and-trade/protocols/usforest/2014/biomass_equations.pdf) 

**For Decay**

* Harmon, M.E., Woodall, C.W., Fasth, B., Sexton, J., & Yatkov, M. (2011) *Differences between standing and downed dead tree wood density reduction factors: A comparison across decay classes and tree species*. Research Paper NRS-15. USDA Forest Service, Northern Research Station, Newtown Square, PA. (See Appendix D) [Link to Paper](https://doi.org/10.2737/NRS-RP-15)

**For standing dead tree decay class**

* USDA Forest Service. (2019). Forest Inventory and Analysis national core field guide volume I: Field data collection procedures for phase 2 plots version 9.0. [Link to Website](https://www.fia.fs.usda.gov/library/field-guides-methods-proc/)


## Background information for surface and ground fuel load calculations 

This suite of functions estimates surface and ground fuel loads (i.e., mass per unit area) from line-intercept transect data. The functions follow the general methodology first described in Stephens (2001):

> "Surface and ground fuel loads were calculated by using appropriate equations developed for Sierra Nevada forests (van Wagtendonk et al. 1996, 1998). Coefficients required to calculate all surface and ground fuel loads were arithmetically weighted by the basal area fraction (percentage of total basal area by species) to produce accurate estimates of fuel loads (Jan van Wagtendonk, personal communication, 1999)."

Brown (1974) used coefficient values generalized for entire regions. Stephens (2001) refined the method described by Brown (1974) by using local overstory data to get species-weighted-average coefficient values. 

Details on how `BerkeleyForestAnalytics` calculates duff/litter, fine, and coarse fuel loads are below. However, note that in all cases we assume the user collected field data following Brown (1974) or a similar method in the **Sierra Nevada**. These functions should not be used for data collected in a different manner or region. Additionally, note that to stay consistent with previous studies, we use both live and dead trees to calculate percent basal area by species. 

### Duff and litter loads 

Duff and litter (or combined duff/litter) are measured as depths at specific points along a sampling transect. Van Wagtendonk (1998) developed regressions for duff, litter, and combined duff/litter loadings as a function of depth for 19 different Sierra Nevada conifer species: 

**Insert table here** 

The plot-level fuel load can be calculated using the following equation: 

\(F_{p} = \frac{\sum(F_{t})}{n}\)

*where* 

  * \(F_{p}\) is the fuel load in \(kg/m^2\) for plot p
  * \(F_{t}\) is the fuel load in \(kg/m^2\) for transect t measured in plot p 
  * \(n\) is the number of total number of transects measured at plot p 
  
<br> 

We can calculate \(F_{t}\) using the following equation: 

\(F_{t} = d_{t}*coef_{p}\)

*where*

  * \(F_{t}\) is the fuel load in \(kg/m^2\) for transect t measured in plot p
  * \(d_{t}\) is the depth of duff, litter, or combined duff/litter in cm measured on transect t. If depth was taken at multiple points along the transect, average them together to calculate \(d_{t}\).
  * \(coef_{p}\) is the species-weighted-average coefficient for plot p
  
<br>

We can calculate \(coef_{p}\) by averaging together the different species-specific coefficients for each tree species contributing fuel to the plot, weighted by their local prevalence. Specifically, we weight each species' coefficient by the proportion of total basal area contributed by that species:

\(coef_{p} = \sum((\frac{BA_{sp,p}}{BA_{total,p}})*coef_{sp})\)

*where* 
  
  * \(coef_{p}\) is the species-weighted-average coefficient for plot p
  * \(coef_{sp}\) is the species-specific coefficient for species sp 
  * \(BA_{sp,p}\) is the basal area occupied by species sp in plot p
  * \(BA_{total,p}\) is the total basal area for plot p
  

*A note on units: the Van Wagtendonk (1998) equations require depths in cm and output fuel loads in \(kg/m^2\). Any unit conversions (for input or outputs) must be done by the user. `BerkeleyForestAnalytics` does the necessary unit conversions for you!*


### Fine fuel loads 

Calculating fuel loads represented by transect counts of 1-hour, 10-hour, and 100-hour fuels is more complicated, but follows the same general idea as described for duff and litter above. The plot-level fuel load can be calculated using the following equation: 

\(W_{c,p} = \frac{\sum(W_{c,t})}{n}\)

*where* 

  * \(W_{c,p}\) is the fuel load for timelag class c (1-hour, 10-hour, or 100-hour) for plot p
  * \(F_{c,t}\) is the fuel load for timelag class c (1-hour, 10-hour, or 100-hour) for transect t measured in plot p 
  * \(n\) is the number of total number of transects measured at plot p 

We can calculate \(W_{c,t}\) using the equation provided by Van Wagtendonk *et al.* (1996):  

\(W_{c,t} = \frac{QMD_{c,p} * SEC_{c,p} * SG_{c,p} * SLP_{t} * k * n_{c,t}}{length_{c,t}}\)  

*where* 

  * \(F_{c,t}\) is the fuel load for timelag class c (1-hour, 10-hour, or 100-hour) for transect t measured in plot p 
  * \(QMD_{c,p}\) is the species-weighted-average quadratic mean diameter for fuels in timelag class c for plot p  
  * \(SEC_{c,p}\) is the species-weighted-average secant of acute angle for fuels in timelag class c for plot p 
  * \(SG_{c,p}\) is the species-weighted-average specific gravity for fuels in timelag class c for plot p
  * \(SLP_{t}\) is the slope correction factor for transect t
  * \(k\) is a constant composed of various unit conversion factors (see table below)
  * \(n_{c,t}\) is the number of timelag class c fuel particles intersecting transect t 
  * \(length_{c,t}\) is the length of transect t sampled for fuels in timelag class c 

<br>

**QMD, SEC, and SG**

QMD, SEC, and SG vary by species and timelag class (see tables below with values from van Wagtendonk (1996)). We can calculate \(QMD_{c,p}\) using the following equation:

\(QMD_{c,p} = \sum(\frac{BA_{sp,p}}{BA_{total,p}}*QMD_{c,sp})\)

*where* 
  
  * \(QMD_{c,p}\) is the species-weighted-average quadratic mean diameter for fuels in timelag class c for plot p
  * \(BA_{sp,p}\) is the basal area occupied by species sp in plot p
  * \(BA_{total,p}\) is the total basal area for plot p
  * \(QMD_{c,sp}\) is the quadratic mean diameter for fuels in timelang class c for species sp 
  
The process is the same for \(SEC_{c,p}\) and \(SG_{c,p})\). 

<br> 

**k values**

Equation constant K for input and output units. These values are from van Wagner (1982) and are used by the US Forest Service per Woodall & Williams (2008). 

Insert table here

